Mixed Precision Euclidean Distance Calculations using GPU Tensor Cores and Applications to Similarity Searches

### 2025-04-13 Paper/main.tex
Scaled diagrams for paper.
Added paper from overleaf because compiles timed out.

%%
%% This is file `sample-sigconf.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `all,proceedings,bibtex,sigconf'.tex)
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigconf.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[sigconf, screen, review]{acmart}

@@

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{none} 
%\setcopyright{acmlicensed}
\copyrightyear{2025}
\acmYear{2025}
\acmDOI{XXXXXXX.XXXXXXX}

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference['ICS 2025']{The 39th ACM International Conference on
  Supercomputing}{June 9--11, 2025}{Salt Lake City, USA}
%%
%%  Uncomment \acmBooktitle if the title of the proceedings is different
%%  from ``Proceedings of ...''!
%%
\acmISBN{000-0-0000-0000-0/00/00}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%custom commands

\usepackage{amsmath,amsfonts}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{multirow}
\usepackage[noend]{algpseudocode}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage[dvipsnames]{xcolor}
\usepackage{forest, tikz}
\usepackage{enumitem}
\usepackage{sidecap}
\usepackage{subfigure}
\usepackage{subcaption}
\usepackage{pgfplots}
\usepackage{tikzscale}
% \usepackage[disable]{todonotes} %disable
\usepackage[]{todonotes} %enable
\usepackage{listings}
\lstset{
language=C++,
basicstyle=\small,
showstringspaces=false,
}

@@

@@

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}
\settopmatter{printfolios=true}
\settopmatter{printacmref=false}
%\pagestyle{plain}

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Mixed Precision Euclidean Distance Calculations using GPU Tensor Cores and Application to Similarity Searches}
\subtitle{\normalsize{ICS 2025 Submission
    \textbf{\#NaN} -- Confidential Draft -- Do NOT Distribute!!}}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Tensor cores (TCs) have been shown to offer speedups over typical CUDA core-based distance similarity self-join problems. However, past work has only investigated using FP64 operands for computation. Modern TCs have the ability to operate on lower precision floating point operands as well. Ampere architecture GPUs can even compute matrix multiplications with half-precision inputs (FP16), while accumulating the results of the computation in single precision (FP32). This mixed-precision configuration offers the highest computational throughput out of all floating-point Tensor core configurations available on the Ampere architecture. To achieve the maximum throughput of 312 teraflops, one needs to carefully design their algorithm for significant hierarchical reuse of data, and maximize memory utilization at every level (global, shared, registers). Previous work utilizes a search index data structure to pare the search space down and reduce the number of computations required. In order to maximize the use of GPU hardware, the problem lends itself to a cohesive brute force strategy. This paper compares our cohesive brute force, mixed-precision Tensor core distance similarity self-join algorithm to a state of the art double precision (FP64) Tensor core algorithm, and a state of the art single precision (FP32) cuda core algorithm. We find that on high-dimensional datasets, the brute force approach outperforms the state of the art by a factor of 4-16, depending on the dataset. Additionally, we quantify the accuracy loss of our reduced precision algorithm to be <0.04\%.
\end{abstract}

@@

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{CUDA, Euclidean Distance, GPU, Mixed Precision Floating Point, Self-Join, Tensor Cores}
%% A "teaser" image appears between the author and affiliation
%% information and the body of the document, and typically spans the
%% page.
% \begin{teaserfigure}
%   \includegraphics[width=\textwidth]{sampleteaser}
%   \caption{Seattle Mariners at Spring Training, 2010.}
%   \Description{Enjoying the baseball game from the third-base
%   seats. Ichiro Suzuki preparing to bat.}
%   \label{fig:teaser}
% \end{teaserfigure}

@@

\mike{From Brian's preliminary results it looks like the time for his algorithm may be less than the time needed to index with GDS-Join!!}
\section{Introduction}\label{sec:intro}
\mike{Define TCs at the beginning}

Euclidean distance calculations are fundamental to numerous applications. For example, in the area of data analytics, they are employed for similarity searches~\cite{gowanlock2023optimization}, outlier detection~\cite{zimek2012survey}, k-nearest neighbor searches~\cite{Samet2008}, and clustering~\cite{bottesch2016speeding}, among other tasks. The Euclidean distance calculation is often a major bottleneck in an application, and given the numerous algorithms that rely on Euclidean distance subroutines, improving the performance of this fundamental operation may have a significant impact on numerous application domains.

In this paper, we propose \ouralg, which is a highly optimized GPU tensor core (TC) algorithm that computes Euclidean distance calculations and which is designed for moderate to high dimensional datasets. We consider comparing all points in an input dataset to each other, which has a quadratic time complexity, and only return those points that are within a search distance, $\epsilon$, of each other. This is a common application scenario, as it is assumed or employed as an optimization technique in many of the data analytics applications described above.

Peak throughput, defined here as the maximum number of Tera Floating Point Operations Per Second (TFLOPS) yields an upper bound on performance, as it is the maximum throughput achievable based on hardware constraints. The Nvidia A100 GPU that we use in our evaluation has a peak throughput using FP32 (single precision floats) on CUDA cores of 19.5 TFLOPS, whereas TCs can achieve a throughput of 312 TFLOPS using mixed precision floats (FP16 operands and FP32 accumulation)~\cite{A100}. Given that the peak throughput of TCs outlined above is a factor 16 higher than CUDA cores, it is clear that if they can be efficiently exploited to compute Euclidean distances, there is immense potential for significant performance improvements over the state-of-the-art methods that only employ CUDA cores.

@@

% [label={{\bf D\arabic*}}]
\begin{enumerate}
\item \bfseries{Memory Bottlenecks --} A major challenge that we identify is that the TCs will be underutilized if the data is not efficiently transferred between each level of the GPU's memory hierarchy, i.e., from global memory, to shared memory, and then into registers for computation on TCs. We find that the coordinates of data points need to be stored and accessed in a manner that maximizes bandwidth utilization and eliminates shared memory bank conflicts during loading and storing.
\item \bfseries{Insufficient Programming Flexibility using the Warp Matrix Multiply Accumulate (WMMA) API --} The WMMA API~\cite{wmmaNvidia} provides programmatic access to TCs where threads in a warp cooperate to execute MMA operations on matrix fragments. These fragments are arrays of registers that represent a subset of a larger matrix. The WMMA API has rigid load/store options that yield the bank conflict problem outlined above, and hide important details about how data flows into TCs.

%\mike{Brian: See (3) below. It seems to me that GeMM has an $O(N^3)$ time complexity, whereas Euclidean distance calculations have $O(|D|^2)$. Therefore one would expect that GeMM would be able to better utilize TCs. But, in our paper, we really have a hidden factor $d$, where the time complexity is $O(d\cdot|D|^2)$. But, in practice $d\ll|D|$, so we ignore $d$. But I know you've said that Euclidean distance calculations take the same work as matrix multiply, so I think I'm missing something here.}
%\brian{I think everything you wrote is correct, I think it must be something small that you are missing. GeMM requires $m*n*k$ computations, and distance requires $d*|D|^2$. In this case though, $d = k$ and $m = n = |D|$. So they are identical computations, with the euclidian distance having the sum of the square values added in at the end (which we can ignore when considering complexity). Also, I adjusted (3), hopefully that helps clear things up.}
\item \bfseries{Insufficient Data Reuse --} General Matrix Multiplication, one of the primary applications of TCs, has a high compute to memory access ratio because of its data reuse, and so the peak throughput achievable on TCs is very high. The highest time complexity component of computing the Euclidean distance between all pairs of points in a dataset can be formulated as a GEMM. In order to ensure the TCs are not idle and waiting for registers to be loaded with data elements, the data reuse has to be carefully designed at all levels including DRAM, L2 Cache, Shared Memory, and registers.

@@

\begin{enumerate}
\item When copying matrix data into the streaming multiprocessors, we optimize the entire data path from global memory into registers. We employ inline Parallel Thread Execution (PTX) instructions that provide the flexibility of non-standard memory access patterns, such that load/store instructions do not cause shared memory bank conflicts.
\item We design a two stage pipeline that includes asynchronous memory copies for loading data from global memory into shared memory and then into registers, such that each warp overlaps TC computation with asynchronous loads. This reduces the duration of memory stalls from global memory, and memory pressure on limited shared memory resources.
\item We optimize storing data, such that each CUDA block stores multiple "k-slices" of matrix data in shared memory, and each warp stores multiple matrix fragments in registers that create data reuse. We also dispatch the work in a specific order to achieve high L2 cache reuse that obviates global memory accesses. 
\item We show that \ouralg can achieve up to XXX\% of peak mixed precision TC performance. We compare \ouralg, to both TC and CUDA core implementations in the literature and demonstrate that it is superior in most experimental scenarios.
\item We demonstrate that when compared with an algorithm using double precision operands, a mixed precision algorithm can produce highly accurate results on real world datasets.
\end{enumerate}

@@

\subsection{Definitions and Problem Statement}\label{sec:problem_statement}
\noindent\textbf{Definition of a dataset and data points --} Let $D$ be a dataset containing $|D|$ points, where each point is defined as $p_i\in D$, and $i=1, 2, \ldots, |D|$. Each point, $p_i$, is defined by coordinates in $d$ dimensions. We define the coordinates for point $p_i$ as $x_j$, where $j=1, 2, \ldots, d$, and therefore, $x_j=\{x_1, x_2, \ldots, x_d\}$.

\noindent\textbf{Definition of the Euclidean Distance --} In what follows, we outline the definition of the Euclidean distance, where we first describe the standard definition and then describe the definition that can be computed on tensor cores. Let $p\in D$ and $q \in D$ be two points in the dataset, $D$, where $p_j$ and $q_j$ refer to the $j^{th}$ coordinates of points $p$ and $q$, respectively. The Euclidean distance between $p$ and $q$ is typically defined as follows:

@@

While this definition is impossible to compute on TCs, a simple expansion yields a form that can be easily computed on TCs: ~\cite{gallet2022leveraging}:

@@

\noindent\textbf{Definition of queries and the self-join --} In this paper we compare all points $p_i \in D$ to each other, which is a high throughput operation that is well-suited to the GPU. In database terminology, this is referred to as a self-join, where conceptually $D$ can be considered a database table, and it is joined on itself. A straightforward implementation of this operation requires three nested for loops.

\subsection{Two Scenarios for Euclidean Distance Calculations}
There are numerous applications that require the calculation of Euclidean distances (Section~\ref{sec:intro}). We describe three common application scenarios below.

@@

\subsubsection{Scenario 2: Index-Supported Distance Similarity Searches}\label{sec:scenario2_index}
To improve the performance of brute force similarity searches (described above), algorithms will typically first construct an indexing data structure that is used to prune the search between points that are sufficiently far from each other in the data space. Then, when a given query point is compared to all other points in the dataset, it only needs to consider a subset of the $|D|$ points, thus reducing the total number of distance comparisons that are required of a brute force search. While our proposed algorithm, \ouralg, does not use an index, we compare it to state-of-the-art GPU implementations do use one.

@@

In this section, we discuss the operation of TCs and how they are programmed on Nvidia GPUs. Tensor Cores can operate on many different kinds of floating point data including FP64, FP32, and FP16-FP32 mixed precision operands. There are several libraries that directly employ TCs for linear algebra operations, including cuBLAS~\cite{cublas} and CUTLASS~\cite{cutlass}. These libraries are suitable when their subroutines can be directly embedded into an algorithm. However, they are unsuitable for our purposes, as they do not offer sufficient granularity and control. There are two other ways to use TCs that provide more flexibility, the WMMA API, and inline PTX instructions.~\cite{wmmaNvidia}. 
\brian{This citation might need PTX instrucitons added too.}

The WMMA API and PTX instructions both move data from shared memory into registers that are shared across the entire warp to compute small submatrices. However, the WMMA API is limited to larger matrix sizes~\cite{tensorcoresizes}, does not specify the register layout, and gives less control over memory addressing. The focus of this paper is on using the PTX instructions as they provide the most flexibility to achieve the highest throughput.
\brian{We probably want to reference the PTX tensor core sizes here to compare WMMA to PTX}

There are two main PTX instructions for using TCs. The first is \texttt{ldmatrix} which creates a fragment of a matrix that is stored in registers shared across a warp. The second is \texttt{mma.sync} which computes $D=A\times B+C$, where $A$, $B$, $C$, and $D$ are each fragments that were created by prior \texttt{ldmatrix} instructions. The $D$ fragment can be substituted for $C$ to accumulate the result of the multiplication in place. In mixed-precision modes, the data stored in matrices $A$ and $B$ are represented in FP16 as $16\times16$ and $16\times8$ fragments while the accumulator matrices, $C$ and $D$, are represented in FP32 as $16\times8$ fragments.

@@

% \begin{table}[!t]
% \caption{Summary of data types and matrix fragment sizes that we consider in this paper. For additional sizes see the CUDA documentation~\cite{tensorcoresizes}. FP16 refers to half precision, FP32 refers to full precision, and FP64 refers to double precision.}
% \mike{Brian double check that these are all the matrix sizes and combinations that we consider in the paper. For example, we don't attempt to accumulate in FP16, right?}
% \brian{You are correct we don't attempt to accumulate in FP16 ever, it is the same throughput supposedly as accumulating in FP32. I am actually only looking at $16\times8\times16$ in my algorithm though. The only other option for inline ptx is 16x8x8. Those other options that are in the table, are only for the WMMA API I believe.}
% \mike{Got it, I'll remove the table.}
% \footnotesize
%     \centering
%         \begin{tabularx}{\columnwidth}{l|l|l|l} 
%         \hline
%         Matrix A&Matrix B&Accumulator Matrix ($C$\&$D$)&Matrix size ($m\times n \times k$)\\\hline
%         FP16&FP16&FP32&16$\times$16$\times$16\\
%         FP16&FP16&FP32&32$\times$8$\times$16\\
%         FP16&FP16&FP32&8$\times$32$\times$16\\
%         FP64&FP64&FP64&8$\times$8$\times$4\\
%         \hline
%         \end{tabularx}
%     \label{tab:matrix_sizes}
% \end{table}

>>>>

%%
%% This is file `sample-sigconf.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `all,proceedings,bibtex,sigconf'.tex)
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigconf.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[sigconf]{acmart}

@@

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmlicensed}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{XXXXXXX.XXXXXXX}
%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[Conference acronym 'XX]{Make sure to enter the correct
  conference title from your rights confirmation email}{June 03--05,
  2018}{Woodstock, NY}

%% These commands are for a PROCEEDINGS abstract or paper.
%%
%%  Uncomment \acmBooktitle if the title of the proceedings is different
%%  from ``Proceedings of ...''!
%%
\acmISBN{978-1-4503-XXXX-X/2018/06}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%custom commands
% For importing pgf plots
\usepackage[utf8]{inputenc}

\usepackage{amsmath,amsfonts}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{multirow}
\usepackage[noend]{algpseudocode}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage[dvipsnames]{xcolor}
\usepackage{forest, tikz}
\usepackage{enumitem}
\usepackage{sidecap}
\usepackage{subfigure}
\usepackage{subcaption}
\usepackage{pgfplots}

\usepackage{tikzscale}
% \usepackage[disable]{todonotes} %disable
\usepackage[]{todonotes} %enable
\usepackage{listings}
\lstset{
language=C++,
basicstyle=\small,
showstringspaces=false,
}

@@

\definecolor{OliveGreen}{cmyk}{0.64,0,0.95,0.40}
\definecolor{CadetBlue}{cmyk}{0.62,0.57,0.23,0}
\definecolor{lightlightgray}{gray}{0.93}
\lstset{
language=C++,                          % Code langugage
basicstyle=\ttfamily,                   % Code font, Examples: \footnotesize, \ttfamily
keywordstyle=\color{OliveGreen},        % Keywords font ('*' = uppercase)
commentstyle=\color{gray},              % Comments font
numbers=left,                           % Line nums position
numberstyle=\tiny,                      % Line-numbers fonts
stepnumber=1,                           % Step between two line-numbers
numbersep=5pt,                          % How far are line-numbers from code
backgroundcolor=\color{lightlightgray}, % Choose background color
frame=none,                             % A frame around the code
tabsize=2,                              % Default tab size
captionpos=t,                           % Caption-position = bottom
breaklines=true,                        % Automatic line breaking?
breakatwhitespace=false,                % Automatic breaks only at whitespace?
showspaces=false,                       % Dont make spaces visible
showtabs=false,                         % Dont make tabls visible
columns=flexible,                       % Column format
morekeywords={__global__, __device__},  % CUDA specific keywords
}

@@

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}
% \settopmatter{printfolios=true}
% \settopmatter{printacmref=false}
%\pagestyle{plain}

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Mixed Precision Euclidean Distance Calculations using GPU Tensor Cores and Applications to Similarity Searches}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.

\renewcommand{\shortauthors}{Trovato et al.}

\begin{abstract}
Modern GPUs are equipped with tensor cores (TCs) that are used for matrix multiplication tasks that are most common in artificial intelligence workloads. TCs have significantly higher computational throughput compared to CUDA cores, so exploiting them for a broader range of algorithms and workloads can lead to significant performance gains. In this paper, we examine using TCs to compute Euclidean distance calculations, which are used in many large-scale data analytics applications. Prior work in this area has only investigated using 64~bit floating point (FP64) operands for computation; however, modern TCs have the ability to operate on lower/mixed precision floating point operands as well (i.e., matrix multiplication in 16 bits and accumulation in 32 bits), which we hereafter refer to as FP16-32. The peak throughput of FP16-32 operations is much higher than for FP64 operations and so these lower precision data types present a major challenge for TC utilization, as it presents a major memory bottleneck, as the TCs are starved for data.
To achieve high computational throughput, we carefully design our algorithm for significant hierarchical reuse of data, and maximize memory utilization at every level (global memory, shared memory, and registers). This paper compares our mixed-precision TC Euclidean distance algorithm to other brute-force methods and also examines the application of similarity searches which typically employ an indexing data structure to eliminate superfluous Euclidean distance calculations. We compare to the state-of-the-art TC Euclidean distance algorithm in the literature that employs FP64, using both brute force and indexing strategies, and a single precision (FP32) CUDA core algorithm that employs an index. We find that across four real-world high-dimensional datasets spanning 128-960 dimensions, the brute force approach achieves a speedup over the state-of-the-art algorithms by a factor of 4--16$\times$, depending on the dataset. Additionally, we quantify the accuracy loss of our mixed precision algorithm to be $<$0.04\% compared to the FP64 baseline.

% similarity self-join algorithm to a state of the art double precision (FP64) Tensor core algorithm, and a state of the art single precision (FP32) cuda core algorithm.

% Ampere architecture GPUs can even compute matrix multiplications with half-precision inputs (FP16), while accumulating the results of the computation in single precision (FP32). This mixed-precision configuration offers the highest computational throughput out of all floating-point Tensor core configurations available on the Ampere architecture.

% Previous work utilizes a search index data structure to pare the search space down and reduce the number of computations required. In order to maximize the use of GPU hardware, the problem lends itself to a cohesive brute force strategy. 
\end{abstract}

@@

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{CUDA, Euclidean Distance, GPU, Mixed Precision Floating Point, Self-Join, Similarity Search, Tensor Cores}
%% A "teaser" image appears between the author and affiliation
%% information and the body of the document, and typically spans the
%% page.
% \begin{teaserfigure}
%   \includegraphics[width=\textwidth]{sampleteaser}
%   \caption{Seattle Mariners at Spring Training, 2010.}
%   \Description{Enjoying the baseball game from the third-base
%   seats. Ichiro Suzuki preparing to bat.}
%   \label{fig:teaser}
% \end{teaserfigure}

@@

\section{A few latex formatting notes (see latex source):}
\begin{itemize}
\item Quotes: Should be formatted as ``my quoted text'' and not "my quoted text"
\item Line breaks and tilde: typically you don't want a reference to a figure, equation, or table number to break over a line. So you should write Table~\ref{tab:WMMA_vs_PTX} and not Table \ref{tab:WMMA_vs_PTX}, i.e., use the tilde to keep the word Table and 1 together.
\end{itemize}

\section{Introduction}\label{sec:intro}
Euclidean distance calculations are fundamental to numerous applications. For example, in the area of data analytics, they are employed for distance similarity searches~\cite{gowanlock2023optimization}, outlier detection~\cite{zimek2012survey}, k-nearest neighbor searches~\cite{Samet2008}, and clustering~\cite{bottesch2016speeding}, among other tasks. The Euclidean distance calculation is often a major bottleneck in an application, and given the numerous algorithms that rely on Euclidean distance subroutines, improving the performance of this fundamental operation can have a significant impact on numerous application domains.

In this paper, we propose \ouralg, which is a highly optimized GPU tensor core (TC) algorithm that computes Euclidean distance calculations and which is designed for moderate to high dimensional datasets\footnote{The code will be made publicly available after paper acceptance.}. We consider comparing all points in an input dataset to each other, which has a quadratic time complexity, and only return those points that are within a search distance, $\epsilon$, of each other. This is a common application scenario, as it is assumed in many of the data analytics applications described above.

Peak throughput, defined here as the maximum number of Tera Floating Point Operations Per Second (TFLOPS) yields an upper bound on performance, as it is the maximum throughput achievable based on hardware constraints. The Nvidia A100 GPU that we use in our evaluation has a peak throughput using FP32 on CUDA cores of 19.5 TFLOPS, whereas TCs can achieve a throughput of 312 TFLOPS using FP16-32~\cite{A100}\footnote{We use CUDA terminology throughout this paper.}. Given that the peak throughput of TCs outlined above is a factor 16 greater than CUDA cores, it is clear that if they can be efficiently exploited to compute Euclidean distances, there is immense potential for significant performance improvements over the state-of-the-art methods that only employ CUDA cores.

@@

% [label={{\bf D\arabic*}}]
\begin{enumerate}
\item \bfseries{Memory Bottlenecks --} A major challenge that we identify is that the TCs will be underutilized if the data is not efficiently transferred between each level of the GPU's memory hierarchy, i.e., from global memory, to shared memory, and then into registers for computation on TCs. We find that the coordinates of data points need to be stored and accessed in a manner that maximizes bandwidth utilization and eliminates shared memory bank conflicts during loading and storing.
\item \bfseries{Insufficient Programming Flexibility using the Warp Matrix Multiply Accumulate (WMMA) API --} The WMMA API~\cite{wmmaNvidia} provides programmatic access to TCs where threads in a warp cooperate to execute MMA operations on matrix fragments. These fragments are arrays of registers that represent a subset of a larger matrix. The WMMA API has rigid load/store options that yield the bank conflict problem outlined above, and hide important details regarding accessing registers and shared memory.

%\mike{Brian: See (3) below. It seems to me that GeMM has an $O(N^3)$ time complexity, whereas Euclidean distance calculations have $O(|D|^2)$. Therefore one would expect that GeMM would be able to better utilize TCs. But, in our paper, we really have a hidden factor $d$, where the time complexity is $O(d\cdot|D|^2)$. But, in practice $d\ll|D|$, so we ignore $d$. But I know you've said that Euclidean distance calculations take the same work as matrix multiply, so I think I'm missing something here.}
%\brian{I think everything you wrote is correct, I think it must be something small that you are missing. GeMM requires $m*n*k$ computations, and distance requires $d*|D|^2$. In this case though, $d = k$ and $m = n = |D|$. So they are identical computations, with the euclidian distance having the sum of the square values added in at the end (which we can ignore when considering complexity). Also, I adjusted (3), hopefully that helps clear things up.}
\item \bfseries{Insufficient Data Reuse --} General Matrix Multiplication (GEMM), one of the primary applications of TCs, has a high compute to memory access ratio because of its data reuse, and so the peak throughput achievable on TCs is very high. The highest time complexity component of computing the Euclidean distance between all pairs of points in a dataset can be formulated as a GEMM. In order to ensure the TCs are not idle and waiting for registers to be loaded with data elements, the data reuse has to be carefully designed at all levels including global memory (DRAM), L2 cache, shared memory, and registers.

@@

\begin{enumerate}
\item When copying matrix data into the streaming multiprocessors (SMs), we optimize the entire data path from global memory into registers. We employ inline Parallel Thread Execution (PTX) instructions that provide the flexibility of non-standard memory access patterns, such that load/store instructions do not cause shared memory bank conflicts.
\item We design a two stage pipeline that includes asynchronous memory copies for loading data from global memory into shared memory and then into registers, such that each warp overlaps TC computation with asynchronous loads. This reduces the duration of memory stalls from global memory, and memory pressure on limited shared memory resources.
\item We optimize storing data, such that each CUDA block stores multiple ``k-slices'' of matrix data in shared memory, and each warp stores multiple matrix fragments in registers that create data reuse. We also dispatch the work in a specific order to achieve high L2 cache reuse that obviates slow global memory accesses. 
\item We show that \ouralg can achieve up to 49\% of peak FP16-32 TC performance. We compare \ouralg to both TC and CUDA core state-of-the-art algorithms and demonstrate that it is superior in most experimental scenarios. Furthermore, our FP16-32 approach yields negligible accuracy loss compared to the FP64 baseline.
\end{enumerate}

@@

\subsection{Definitions and Problem Statement}\label{sec:problem_statement}
\noindent\textbf{Definition of a dataset and data points --} Let $D$ be a dataset containing $|D|$ points, where each point is defined as $p_i\in D$, and where $i=1, 2, \ldots, |D|$. Each point, $p_i$, is defined by coordinates in $d$ dimensions. We define the coordinates for point $p_i$ as $x_j$, where $j=1, 2, \ldots, d$, and therefore, $x_j=\{x_1, x_2, \ldots, x_d\}$.

\noindent\textbf{Definition of the Euclidean Distance --} In what follows, we outline the definition of the Euclidean distance, where we describe the standard definition and then describe the definition that can be computed on TCs. Let $p\in D$ and $q \in D$ be two points in the dataset, $D$, where $p_j$ and $q_j$ refer to the $j^{th}$ coordinates of points $p$ and $q$, respectively. The Euclidean distance between $p$ and $q$ is typically defined as follows:

@@

While this definition is impossible to compute on TCs, a simple expansion yields a form that can be computed on TCs: ~\cite{gallet2022leveraging}:

@@

\noindent\textbf{Definition of queries and the distance similarity self-join --} In this paper we compare all points $p_i \in D$ to each other, which is a high throughput operation that is well-suited to the GPU. In database terminology, this is referred to as a self-join, where conceptually $D$ can be considered a database table, and it is joined on itself. A straightforward implementation of this operation requires three nested for loops (two for comparing each point to every other point, and one for the distance calculation between each pair of points).

\subsection{Scenarios for Euclidean Distance Calculations}
There are numerous applications that require the calculation of Euclidean distances (Section~\ref{sec:intro}). We describe two common application scenarios below.

@@

\subsubsection{Scenario 2: Index-Supported Distance Similarity Searches}\label{sec:scenario2_index}
To improve the performance of brute force similarity searches (described above), algorithms will typically first construct an indexing data structure that is used to prune the search between points that are sufficiently far from each other in the data space. Then, when a given query point is compared to all other points in the dataset, it only needs to consider a subset of the $|D|$ points, thus reducing the total number of distance comparisons that are required of a brute force search. While our proposed algorithm, \ouralg, does not use an index, we compare it to GPU algorithms that use an index.

@@

In this section, we discuss the operation of TCs and how they are programmed on Nvidia GPUs. TCs can operate on different floating point data types including FP64, FP32, and FP16-32. There are several libraries that directly employ TCs for linear algebra operations, including cuBLAS~\cite{cublas} and CUTLASS~\cite{cutlass}. These libraries are suitable when their subroutines can be directly embedded into an algorithm. However, they are unsuitable for our purposes, as they do not offer sufficient granularity and control. There are two other ways to use TCs that provide more flexibility, the WMMA API, and inline PTX instructions~\cite{wmmaNvidia, ptxNvidia}.

The WMMA API and PTX instructions both load data from shared memory into registers that are shared across an entire warp to compute small submatrices. However, the WMMA API is limited to larger matrix sizes (see Table~\ref{tab:WMMA_vs_PTX} for a comparison), does not specify the register layout, and yields less control over memory addressing. In this paper, we use PTX instructions as they provide the most flexibility to achieve the highest throughput.

\begin{table}[!t]
\caption{Comparison of WMMA and PTX matrix sizes for FP16-32 (16~bit matrix multiplication and 32~bit accumulation)~\cite{tensorcoresizes, ptxNvidia}.}
\footnotesize
    \centering
        \begin{tabularx}{\columnwidth}{|L|l|l|} 
        \hline
        Size ($m$-$n$-$k$)&WMMA&PTX\\\hline
        16$\times$16$\times$16&\checkmark&\\
        32$\times$8$\times$16&\checkmark&\\
        8$\times$32$\times$16&\checkmark&\\
        \hline
        \end{tabularx}
    \label{tab:WMMA_vs_PTX}

\brian{We probably want to reference the PTX tensor core sizes here to compare WMMA to PTX}
\mike{Agreed. Can you update the table with the matrix sizes for PTX from here? I don't want to make an error. Link commented in source below}
% https://docs.nvidia.com/cuda/parallel-thread-execution/#asynchronous-multiply-and-accumulate-instruction-wgmma-mma-async    
\end{table}

There are two main PTX instructions for using TCs. The first is \texttt{ldmatrix} which creates a fragment of a matrix that is stored in registers shared across a warp. The second is \texttt{mma.sync} which computes $D=A\times B+C$, where $A$, $B$, $C$, and $D$ are each fragments that were created by prior \texttt{ldmatrix} instructions. The $D$ fragment can be substituted for $C$ to accumulate the result of the multiplication in place. Using FP16-32 instructions, the data stored in matrices $A$ and $B$ are represented in FP16 as 16$\times$16 and 16$\times$8 fragments while the accumulator matrices, $C$ and $D$, are represented in FP32 as 16$\times$8 fragments.

@@

% \begin{table}[!t]
% \caption{Summary of data types and matrix fragment sizes that we consider in this paper. For additional sizes see the CUDA documentation~\cite{tensorcoresizes}. FP16 refers to half precision, FP32 refers to full precision, and FP64 refers to double precision.}
% \mike{Brian double check that these are all the matrix sizes and combinations that we consider in the paper. For example, we don't attempt to accumulate in FP16, right?}
% \brian{You are correct we don't attempt to accumulate in FP16 ever, it is the same throughput supposedly as accumulating in FP32. I am actually only looking at 16$\times$8$\times$16 in my algorithm though. The only other option for inline ptx is 16x8x8. Those other options that are in the table, are only for the WMMA API I believe.}
% \mike{Got it, I'll remove the table.}
% \footnotesize
%     \centering
%         \begin{tabularx}{\columnwidth}{l|l|l|l} 
%         \hline
%         Matrix A&Matrix B&Accumulator Matrix ($C$\&$D$)&Matrix size ($m\times n \times k$)\\\hline
%         FP16&FP16&FP32&16$\times$16$\times$16\\
%         FP16&FP16&FP32&32$\times$8$\times$16\\
%         FP16&FP16&FP32&8$\times$32$\times$16\\
%         FP64&FP64&FP64&8$\times$8$\times$4\\
%         \hline
%         \end{tabularx}
%     \label{tab:matrix_sizes}
% \end{table}

\subsection{Literature on Expanding TC Applications}
There are few papers in the literature that extend GPU TCs to address algorithms and application scenarios beyond matrix multiplication and accumulation operations. Most of the papers below are not directly related to this paper, as they examine differing use cases for TCs; however, we summarize these developments to give the reader a broader picture of the burgeoning utility of GPU TCs.
